---
title: "Data622_HW3"
author: "Mengqin Cai"
date: "4/5/2021"
output: 
  html_document:
    df_print: paged
    toc: yes
    toc_collapsed: yes
    toc_float: yes
  md_document:
    variant: markdown_github
  pdf_document:
        extra_dependencies: ["geometry", "multicol", "multirow"]
theme: lumen
number_sections: yes
toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,include=FALSE}
library(palmerpenguins)
library(dplyr)
library(ggplot2)
data(penguins)
library(caret)
library(pROC)
library(GGally)
library(tidyr)
library(Amelia)
library(VIM)
library(MASS)
library(psych)
library(class)
library(tree)
library(rpart)
library(rpart.plot)
library(randomForest)

```



# Problem 1: KNN

Please use K-nearest neighbor (KNN) algorithm to predict the species variable. Please be sure to walk through the steps you took. (40 points)

```{r }
head(penguins)
summary(penguins)

```

To better evaluate the model, I split the dataset into training and test set. 


First, check the missing value of the whole dataset and use KNN imputation to impute the dataset

```{r}
aggr(penguins,bars=T, numbers=T, sortVars=T)
penguins<-kNN(penguins)
penguins<-subset(penguins,select=species:sex)
missmap(penguins)
```

```{r}

levels(penguins$species) <- c("Adelie", "Chinstrap", "Gentoo")
penguins$species<-as.numeric(penguins$species)

levels(penguins$island) <- c("Biscoe", "Dream", "Torgersen")
penguins$island<-as.numeric(penguins$island)

levels(penguins$sex) <- c("female", "male")
penguins$sex<-as.numeric(penguins$sex)
```


```{r}
normalize <- function(x) {
return ((x - min(x)) / (max(x) - min(x))) }
head(penguins)
penguins_Trans<- as.data.frame(lapply(penguins[,3:6], normalize))
penguins_Trans<-cbind(penguins[,1],penguins[,2],penguins_Trans,penguins[,7] )
colnames(penguins_Trans)<-c("species","island","bill_length_mm","bill_depth_mm","flipper_length_mm","body_mass_g","sex")
head(penguins_Trans)
```





```{r}
sample_size<-floor(0.8*nrow(penguins))
set.seed(123)
train_ind<-sample(seq_len(nrow(penguins)),size=sample_size)
train_penguins<-penguins_Trans[train_ind,]
test_penguins<-penguins_Trans[-train_ind,]
```





Fit the model
```{r}
set.seed(123) 
sqrt(nrow(train_penguins))
k16<-knn(train_penguins,test_penguins,cl=train_penguins$species,k=16)
k17<-knn(train_penguins,test_penguins,cl=train_penguins$species,k=17)
```
```{r}
misClassError <- mean(k16 != test_penguins$species)
misClassError
table(k16,test_penguins$species)


misClassError <- mean(k17 != test_penguins$species)
misClassError
table(k17,test_penguins$species)
```
There is no different with K16 or K17, we can choose either of the model.

# Problem 2: Decision Trees

Please use the attached dataset on loan approval status to predict loan approval using Decision Trees. Please be sure to conduct a thorough exploratory analysis to start thetask and walk us through your reasoning behind all the steps you are taking. 



```{r}
loan<-read.csv("https://raw.githubusercontent.com/DaisyCai2019/NewData/master/Loan_approval.csv")
head(loan)
summary(loan)
```
```{r}
missmap(loan)
loanTrans<-kNN(loan)%>%
          subset(select = Loan_ID:Loan_Status)
```

```{r}

loanTrans$Loan_Status<-factor(loanTrans$Loan_Status)


loanTrans<-loanTrans %>%
           mutate(Gender = factor(Gender),
                  Married = factor(Married),
                  Dependents=factor(Dependents),
                  Education=factor(Education),
                  Self_Employed=factor(Self_Employed),
                  Property_Area=factor(Property_Area),
                  Loan_Status=factor(Loan_Status))

summary(loanTrans)
```


```{r}
set.seed(123)
train_sample<-sample(1:nrow(loanTrans),size = floor(0.80*nrow(loanTrans)))
train_loan<-loanTrans[train_sample,]
test_loan<-loanTrans[-train_sample,]
```


```{r}
tree<- rpart(Loan_Status~Gender+Married+Dependents+Education+Self_Employed+ApplicantIncome+CoapplicantIncome+LoanAmount+Loan_Amount_Term+Credit_History+Property_Area,data=train_loan)
rpart.plot(tree,nn=TRUE)
summary(tree)

```
```{r}
loanPre<-predict(tree,test_loan,type="class")
table(loanPre,test_loan$Loan_Status)
accuracy<-mean(loanPre==test_loan$Loan_Status)
accuracy
```


# Problem 3: Random Forests


Using the same dataset on Loan Approval Status, please use Random Forests to predict on loan approval status. Again, please be sure to walk us through the steps you took to get to your final model. (50 points)



```{r}
rf <- randomForest(Loan_Status~Gender+Married+Dependents+Education+Self_Employed+ApplicantIncome+CoapplicantIncome+LoanAmount+Loan_Amount_Term+Credit_History+Property_Area,data=train_loan)
rf
importance(rf)
```
```{r}
rfPre<-predict(rf,test_loan)
table(rfPre,test_loan$Loan_Status)
accuracy2<-mean(rfPre==test_loan$Loan_Status)
accuracy2
```


# Problem 4: Gradient Boosting

Using the Loan Approval Status data, please use Gradient Boosting to predict on the loan approval status. Please use whatever boosting approach you deem appropriate;but please be sure to walk us through your steps. (50 points)











# Problem 5: Gradient Boosting

Model performance: please compare the models you settled on for problem # 2 – 4.Comment on their relative performance. Which one would you prefer the most? Why?(20 points)





